
# Survival Analysis

It is also called _time-to-event_ analysis

events:
  - patient events/outcome (death or [nonfatal outcomes] hospital readmission, etc)
  - study participants 
      - lost to follow-up in a study
      - censored = participant dropped out of study
      
methods:
  - logistic regression (binary events). Interested in whether outcome happens (when does not matter)
  - survival analysis interested in if outcome happened, how long it took for that outcome to occur {time to event}
  

if developed a disease, what statistic would determine how long to live and chances of surviving
the Kaplan-Meier plot and Log-rank test answer these

TNM staging system 
t = size of tumor, lymph nodes involved
m = describes distant metasis spread to other body parts 


## Life tables

Life tables are used to measure the probability of death at a given age and the life expectancy at varying ages.


types:

- cohort (generational) life table
- current or period life table

Cohort life tables take an actual set of people born at the same time, usually in the same year or even on the same day of the same year, and follow them up for their whole lives.

Period life tables take a hypothetical cohort of people born at the same time and uses the assumption that they are subject to the age-specific mortality rates of a region or country. These rates are often calculated using census data as the base population and actual age-specific death rates during the census year (and typically also one year either side).


| Time (t) days     | patients alive at _t_ | patients who died at  _t_ | Probability: past time _t_ |
|-------------------|-----------------------|---------------------------|----------------------------|
| 0 (study start)   |   100                 | 0                         |       1                    |
| 1                 |   100                 | ?                         |   ?                        |

people drop out of studies or 'lost to follow up' (*censored*)



Calculate a Kaplan-Meier table

The plot of the survival function versus time is called the survival curve.

The Kaplan-Meier method can be used to estimate this curve from the observed survival times without the assumption of an underlying probability distribution.

Example [Survival Analysis](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1065034/)

```{r}
table1 = data.frame(
  patient_number = 1:15,
  survival_time_days = c(1,1,4,5,6,8,9,9,12,15,22,25,37,55,72),
  outcome = c('died','died','died','died','unknown',
              'died','survived','died','died','unknown',
              'died','survived','died','died','survived'),
  treatment = c(2,2,2,2,2,1,2,2,1,1,2,1,1,1,1),
  age_yrs = c(75,79,85,76,66,75,72,70,71,73,66,73,68,59,61)
)

table1

```


| time (t) days | event                     |
|---------------|---------------------------|
| 0 start       | 8 patients recruited      |
| 1             | 2 patients die            |
| 4             | 1 patient dies            |
| 5             | 1 patient dies            |
etc

compute the probability of survival past time t,
the probability of surviving past day t is simply the probability of surviving past day t-1 times the proportion of patients that survive on day t.

| Time (t) days     | alive at _t_ | died at  _t_ | Proportion: past time _t_  | Probability: past _t_|
|-------------------|--------------|--------------|----------------------------|----------------------|
| 0 (study start)   | 8            | 0            |  (8-0)/8 = 1               | 1                    |
| 1                 | 8            | 2            | (8-2)/8 = 0.75             | 1 * 0.75             |
| 4                 | 6            | 1            | (6-1)/6 = 0.83             | 0.75 * 0.83 = 0.623  |
| 5                 | 5            | 1            | (5-1)/5 = 0.80             | 0.623 * 0.80 = 0.498 |


```{r}
library(tidyverse)

df = data.frame(
  probability_of_survival = c(1, 0.75, 0.623, 0.498),
  survival_time_days = c(0,1,4,5)
)

ggplot2::ggplot(df, aes(x=survival_time_days, y= probability_of_survival))+geom_point()

```



if you connect the dots it would result in steps





## data

The data are simulated based on real hospital administrative data for England called Hospital Episodes Statistics. Your simulated extract contains a random sample of emergency (unplanned) admissions for heart failure (ICD10 code I50).

1 indicates that the patient had it recorded

data dictionary :

- death
- los (hospital length of stay)
- age (years)
- gender binary (1=m, 2=f)
- cancer
- cabg
- crt (cardiac resynchronization device)
- defib (implant)
- dementia
- diabetes (all types)
- hyptertension
- ihd (ischemic heart disease)
- mental health
- arrhythmias
- copd
- obesity
- pvd
- renal idease
- vulvular disease (heart valves)
- metastatic cancer
- pacemaker
- prior appts attented (outpatient appointments attended in previous year)
- prior dnas (outpatient appointments misses in previous year)
- pci percutaneous coronary intervention
- stroke
- senile
- quintile (socioeconomic status: 1 (affluent) to 5 (poor))
- ethnic group (1=white, 2=black, 3= Indian subcontinent, 8= unknown, 9= other)
- fu time (follow up time)



```{r}
library(tidyverse)
library(survival)

df = read_csv('./COURSERA-data.csv')

glimpse(df)
```


[survival documentation](https://cran.r-project.org/web/packages/survival/survival.pdf)

```{r}

df = df %>% 
  select(gender, fu_time, death) %>% 
  mutate(gender = factor(gender))

# Kaplan-Meier plot
km_fit = survival::survfit(Surv(fu_time, death) ~ 1)

plot(km_fit)
```


“Survfit” fits a simple survival model that depends only on gender in terms of predictors
(in this case, 0)
`death= 1` for people who had died by the end of the follow-up period
`death= 0` for those still alive at that time

Technically, those people still alive are censored, because we don’t know when they’ll die

`survfit()` function produces the Kaplan-Meier estimates of the probability of survival over time 


```{r}
summary( km_fit, times = c(1:7, 30, 60, 90 * (1:10)))

# The “times” argument gives us control over what time periods we want to see
# week 1
# 30 days
# 90 days
# every 90 days
```


split the curv by gender (binary)

```{r}
km_gender_fit = survfit(Surv(fu_time, death) ~ gender)

plot(km_gender_fit)
```

To compare survival by gender, we can run a logrank test

gender 1 = male, 2 = female
```{r}
survdiff(Surv(fu_time, death) ~ gender, rho = 0)
```

similar survival rates over time between genders


The log-rank test compares the survival time by gender. It’s the most popular method of comparing the survival of patient groups, you don’t need to know anything about the shape of the survival curve or the distribution of survival times.

It’s based on a comparison of the observed numbers of deaths and the numbers of deaths expected if in fact there were no difference in the probability of death between the groups (genders in this case) and uses a chi-squared test. 

The resulting p-value that you should have got is high, at 0.8. There’s therefore no good evidence of a difference between the genders in their survival times. 
















