

# Fitting Regression Models

in the COPD dataset

walking ability has variables `MWT1, MWT2, MWT1Best` which means 6 minute walking test, distance measured in meters in 6 minute duration

look at the data, understand it, what types are the variables?

```{r}
library(tidyverse)

COPD = read_csv('./COURSERA-copd.csv')

glimpse(COPD)
```

look at the data

```{r}
hist(COPD$MWT1Best, breaks = 12)
```

the distribution looks symmetric, peaks between 400 and 500, have an outlier past 650


`subset(COPD, COPD$MWT1Best > 650)` is the same as filter()

```{r}
COPD %>% 
  filter(MWT1Best > 650)
```


look at another variable

```{r}
COPD %>% 
  ggplot( aes(x= FEV1))+
  geom_histogram(bins = 9)
```


Descriptive statistics and two-way plots can tell you more about the data. Useful statistics to summarise your data are: the mean, the standard deviation, the range, the median, and the inter-quartile range.

```{r}

skimr::skim(COPD$MWT1Best)
```


```{r}
skimr::skim(COPD$FEV1)
```


```{r}
COPD %>% 
  select(FEV1, MWT1Best) %>% 
  ggplot(
    aes(x= FEV1, y= MWT1Best)
  )+
  geom_point()+
  labs(x='FEV1', y='MWT1Best')+
  ggdark::dark_theme_gray()
```
the scatterplot shows an even distribution, no outliers, so proceed with Pearson's correlation coefficient to assess a linear association between the two variables. But we will also have a look at the Spearman’s for comparison. 

```{r}
cor.test( COPD$FEV1, COPD$MWT1Best, use="complete.obs", method = 'pearson')
```

```{r}
cor.test(COPD$FEV1, COPD$MWT1Best, use='complete.obs', method = 'spearman')
```


Pearson correlation = 0.4692142 
        95% confidence interval =  between 0.3 to 0.6
        p value is less than 0.001
       strong evidence against the null 
        
        
Spearman has similar results
















